Applying WaveNet on frequency data

### 2017-01-20 paper/paper.tex
added pdf, fixed typos

\section{Introduction}
In 2016, \cite{wavenet} introduced WaveNet, a generative model for raw audio achieving astonishing results in learning sound features and generating harmonious audio. Trained on piano music, highly dynamic piano music varying smoothly in melody, volume and speed could be generated.\\
The aim of this paper is to enhance the WaveNet architecture and apply it to a more feature-oriented representation of input data. Instead of performing a classification task on raw audio, the network is adopted to operate on the frequencies of the audio data, processed by Fouier transformation.\\
As training data in the experiments, piano samples from the MusicNet dataset, a collection of 330 freely-licenced recordings of classical music are taken. The music is recorded with 44100 samples per second. The data is publicly available online \cite{musicnet}.\\
In particular, the following topics are covered:
\begin{itemize}
\item Based on WaveNet from \cite{wavenet}, a network architecture to generate music is introduced operating on the frequencies of audio input.
\item This new network is trained and evaluated on piano music from the MusicNet dataset. Furthermore, optimization approaches as data normalization and variations in preprocessing the data are included in the experiments.
\item Results from the evaluation are analyzed in detail. In the context of the introduced network, advantages as well as weaknesses are discussed. 
\end{itemize}
The source code of the experiments can be found on GitHub.\footnote{\url{https://github.com/sennluzi/tensorflow-wavenet}}

\section{Models and Methods}
The following section describes the underlying model of the network used in the experiments of this paper. As core network the WaveNet architecture with a stack of dilated causal convolutional layers is taken. Hereupon, the data input layer is changed conceptually.
\begin{figure}[tbp]
	\centering
	\includegraphics[width=\columnwidth]{figures/wavenet_dilated_cnn.png}
	\caption{dilated convolutional network}
	\label{fig:wavenet_dilated_cnn}
\end{figure}
\subsection{WaveNet architecture}
Given a sequence of amplitudes, WaveNet predicts the subsequent amplitude. This task is done in a classification setting: The amplitudes are quantized in 256 classes \cite{itu1988711}. The input of the WaveNet network is a sequence of such quantizied amplitudes, each encoded as a 256-dimensional one-hot vector. Performing a softmax-function in its last layer, the network outputs a probability distribution $P(x_{i+1} | x_1, ..., x_i)$ in the $i$th step. The amplitude $x_{i+1}$ of the next step is then sampled according to that distribution.\\
To model the the dependency of an audio sample on previous samples, WaveNet uses a stack of dilated convolutions. Figure \ref{fig:wavenet_dilated_cnn} depicts this architecture exemplarily for such a network of depth 4. By using exponentially growing dilations in depth, a network with $n$ dilated convolutional layers can observe the last $2^n$ input values to predict an output.\\
Training WaveNet is done similar as an autoencoder: The $i$th amplitude of the input data is predicted by only using previous input data and taken as label for the loss calculation and backpropagation. Thus, training only requires the input data and can be done in parallel. Against that, the generation of a audio sample has to be processed sequentially, since the $i$th step relies on the output of the $(i-1)$th one.\\
As a basis for WaveNet, the implementation of Igor Babuschkin is taken \cite{Babuschkin2016}. The original WaveNet paper mentions only very few coefficients for how their network is configured. Therefore reproducing the paper exactly is very hard. The following numbers are assumptions mainly based on the experience of Babuschkin's implementation. In the following experiments, the dilations are $2^n$ for $n$ taking the values $(0,1,...,8)$ and are repeated 5 times. This leads to 50 hidden dilation layers. The convolution filter size is set to $2$ and the number of channels of the residual and the dilation is $32$.\\
\subsection{Data preprocessing}
To extract frequencies, the raw audio data consisting of pressure measurements are divided into chunks of size 2048, which are then transformed into the frequency space using FFT. Therefrom, the first 150 coefficients are taken as dominant frequencies for the subsequent training. To ensure that audio features are transformed smoothly, two neighboring raw audio chunks overlap in a few samples, described by a variable $stride$. Equation \ref{eq:fft} summarized the frequency extraction of the $i^\textnormal{th}$ chunk compactly.
\begin{multline}\label{eq:fft}
\textnormal{frequencies\_coeffs}_i =
FFT(\textnormal{inputData}[i \cdot stride:\\i \cdot stride+2048])[:150]
\end{multline}
Considering real and imaginary parts of the complex Fourier coefficients separately, each chunk of 2048 raw audio samples is transformed into 300 real frequency values.\\
To decrease the dimensionality even further, principle component analysis (PCA) is used to reduce these 300 frequencies to 100 values. These 100 PCA coefficient are considered one tone and fed into the network as one time step.\\
Whereas in the original WaveNet architecture, 16000 samples represented on second of audio, the adopted network only needs to process $44100/2048 \approx 22$ samples per second. However, this change infers a fundamentally different approach of audio training and generation, described in the next section. \\

\subsection{Evaluation of the preprocessing}
The following analyses the amount of information which is lost by the compression steps of the previous session. Figure \ref{fig:pca_explained_variance} plots the explained variance of the first 250 eigenvectors during the The vast majority of the variance of the frequencies is captured by the first 100 PCA components. Thus, PCA is a reasonable measure to reduce the 300 dimensional data from the FFT to 100 PCA components.\\
The best way to evaluate a compression procedure is to compare a compressed and afterwards reconstructed input to the original data. Figure \ref{fig:preprocessing_difference_waveform} compares exemplarily the recorded raw audio input of a music sample from the MusicNet dataset to a compressed and afterwards reconstructed version of the signal. The plot shows that the two signals are similar in their general behavour and only vary vary only in small details.\\
The same results can be observed by listening to the original and the compressed and reconstructed audio file: After preprocessing the audio contains some cracking noise, but nevertheless is the melody and the tone of the music well audible.\\
Summarizing, in neither by cutting after the first 150 FFT coefficients, nor by the PCA sigifican information of the original input is lost. Therefore, the described preprocessing leads to a decent compression of the audio signal.\\

\begin{figure}[htbp]
  \centering
  \includegraphics[width=\columnwidth]{figures/pca_explained_variance}
  \caption{Explained variance of the i'th eigenvector in the principal component analysis (with i being the horizontal axis)}
  \label{fig:pca_explained_variance}
\end{figure}

@@

Nevertheless, the experiments of this paper can be analysed more detailled. Varying the stride and including normalization yielded conceptually different results.\\

@@

A second conceptual attempt during the experiments was to normalize the PCA results before the training. The aim of this additional preprocessing task was to balance the contribution of all 100 PCA dimensions to the loss, hoping that all frequencies are learned more equally. In a first attempt, data normalzed to zero mean and unit variance turned out to produce too small loss for training. With vanishing gradients the network could not learn at all. Therefore, in a second attempt the input data was normalized to zero mean and an equal variance of all dimensions of 10. With this specifications, the network was able to learn as well as without normalization.\\

@@

\section{Discussion}
The approach of this paper could be promising due to the following reason: The original WaveNet paper's network has to learn the behaviour of sound consisting of repetitive oscillations. With transforming the signal to the Fourier space, we take this burden from the neural network. The network can focus on data which is more related to the final tone. For example while a tone is the same for a certain time, the output of the network is also the same.\\
But it might be exactly this phenomenon which makes good music generation impossible. During training, the network might do really well only staying on one and the same pitch, because changes happen in too rare occasions to influence the overall loss.\\
Future work based on that paper, one could try to include the rate of changing tones into the loss function and to punish a network when producing too many times in a row a similar output.\\
There are two reasons one could think of, why this approach's result does not match its expectations:
\begin{itemize}
  \item The original paper quantizes the amplitudes and can then use a softmax layer as a last layer of the network. In this new approach, we need continous values (PCA coefficients). Since the results are much worse than with WaveNet, the missing quantization (and connected to it the missing softmax layer) might be the reason behind this phenomenon.
  \item The original WaveNet has a probability distribution as output. The exact amplitudes are then sampled according to that distribution. In the approach presented in this paper, one relies on continuous values, which are directly calculated (no probabilities included). In future work, one can change the model to produce a probability distribution of the output values.
\end{itemize}

It has been shown that the stride is of great importance for the strength of the frequencies. Strong frequencies are interpreted as the network being confident in its prediction while weak frequencies are a sign of uncertainty. When the training timesteps do not overlap (stride = window size), the network produces much weaker frequencies. But a big stride might also lead to what we experience as not-changing frequencies. With a small stride most of the data chunks are the same as the previous one (when the length of a tone is much longer than one time unit). With larger strides, the frequencies change abruptly from one window to another. When a change in frequencies occurrs on an dataset with smaller strides, the absolute value of the difference of frequencies is much smaller.\\
Therefore one can assume that with changing the setting (stride, but also window size, dilations, normalization etc.), also this approach might produce good results.

\section{Summary}
With WaveNet a very promising music generation network is presented. It relies on predicting amplitudes when the preceeding amplitudes are given. In this paper it is investigated whether transforming the data to the Fourier space improves the predictions. After transforming, the dimensionality of the audio file is vastly diminished. But at the same time, the ability to quantize the network's outputs is lost. Unfortunately the network could not return as good results as the original WaveNet paper. With the adopted network architecture, the generated audio files rapidly converge to one steady tone, while changes of pitches is not learned. However, by the approaches and discussions in the paper yiels conceptual insights and motivate further investigations.

>>>>

\section{Introduction}
In 2016, \cite{wavenet} introduced WaveNet, a generative model for raw audio achieving astonishing results in learning sound features and generating harmonious audio. Trained on piano music, highly dynamic piano music varying smoothly in melody, volume and speed could be generated.\\
The aim of this paper is to enhance the WaveNet architecture and apply it to a more feature-oriented representation of input data. Instead of performing a classification task on raw audio, the network is adopted to operate on the frequencies of the audio data, processed by Fourier transformation.\\
As training data in the experiments, piano samples from the MusicNet dataset, a collection of 330 freely-licensed recordings of classical music are taken. The music is recorded with 44100 samples per second. The data is publicly available on the Internet \cite{musicnet}.\\
In particular, the following topics are covered:
\begin{itemize}
\item Based on WaveNet from \cite{wavenet}, a network architecture to generate music is introduced operating on the frequencies of audio input.
\item This new network is trained and evaluated on piano music from the MusicNet dataset. Furthermore, optimization approaches as data normalization and variations in preprocessing the data are included in the experiments.
\item Results from the evaluation are analyzed in detail. In the context of the introduced network, advantages as well as weaknesses are discussed. 
\end{itemize}
The source code of the experiments can be found on GitHub.\footnote{\url{https://github.com/sennluzi/tensorflow-wavenet}}

\section{Models and Methods}
The following section describes the underlying model of the network used in the experiments of this paper. As core network the WaveNet architecture with a stack of dilated causal convolutional layers is taken. Hereupon, the data input layer is changed conceptually.
\begin{figure}[tbp]
	\centering
	\includegraphics[width=\columnwidth]{figures/wavenet_dilated_cnn.png}
	\caption{dilated convolutional network}
	\label{fig:wavenet_dilated_cnn}
\end{figure}
\subsection{WaveNet architecture}
Given a sequence of amplitudes, WaveNet predicts the subsequent amplitude. This task is done in a classification setting: The amplitudes are quantized in 256 classes \cite{itu1988711}. The input of the WaveNet network is a sequence of such quantizied amplitudes, each encoded as a 256-dimensional one-hot vector. Performing a softmax-function in its last layer, the network outputs a probability distribution $P(x_{i+1} | x_1, ..., x_i)$ in the $i$th step. The amplitude $x_{i+1}$ of the next step is then sampled according to that distribution.\\
To model the the dependency of an audio sample on previous samples, WaveNet uses a stack of dilated convolutions. Figure \ref{fig:wavenet_dilated_cnn} depicts this architecture exemplarily for such a network of depth 4. By using exponentially growing dilations in depth, a network with $n$ dilated convolutional layers can observe the last $2^n$ input values to predict an output.\\
Training WaveNet is done similar as an auto-encoder: The $i$th amplitude of the input data is predicted by only using previous input data and taken as label for the loss calculation and backpropagation. Thus, training only requires the input data and can be done in parallel. Against that, the generation of a audio sample has to be processed sequentially, since the $i$th step relies on the output of the $(i-1)$th one.\\
As a basis for WaveNet, the implementation of Igor Babuschkin is taken \cite{Babuschkin2016}. The original WaveNet paper mentions only very few coefficients for how their network is configured. Therefore reproducing the paper exactly is very hard. The following numbers are assumptions mainly based on the experience of Babuschkin's implementation. In the following experiments, the dilations are $2^n$ for $n$ taking the values $(0,1,...,8)$ and are repeated 5 times. This leads to 50 hidden dilation layers. The convolution filter size is set to $2$ and the number of channels of the residual and the dilation is $32$.\\
\subsection{Data preprocessing}
To extract frequencies, the raw audio data consisting of pressure measurements are divided into chunks of size 2048, which are then transformed into the frequency space using FFT. Therefrom, the first 150 coefficients are taken as dominant frequencies for the subsequent training. To ensure that audio features are transformed smoothly, two neighboring raw audio chunks overlap in a few samples, described by a variable $stride$. Equation \ref{eq:fft} summarized the frequency extraction of the $i^\textnormal{th}$ chunk compactly.
\begin{multline}\label{eq:fft}
\textnormal{frequencies\_coeffs}_i =
FFT(\textnormal{inputData}[i \cdot stride:\\i \cdot stride+2048])[:150]
\end{multline}
Considering real and imaginary parts of the complex Fourier coefficients separately, each chunk of 2048 raw audio samples is transformed into 300 real frequency values.\\
To decrease the dimensionality even further, principle component analysis (PCA) is used to reduce these 300 frequencies to 100 values. These 100 PCA coefficient are considered one tone and fed into the network as one time step.\\
Whereas in the original WaveNet architecture, 16000 samples represented on second of audio, the adopted network only needs to process $44100/2048 \approx 22$ samples per second. However, this change infers a fundamentally different approach of audio training and generation, described in the next section. \\

\subsection{Evaluation of the preprocessing}
The following analyses the amount of information which is lost by the compression steps of the previous session. Figure \ref{fig:pca_explained_variance} plots the explained variance of the first 250 eigenvectors during the The vast majority of the variance of the frequencies is captured by the first 100 PCA components. Thus, PCA is a reasonable measure to reduce the 300 dimensional data from the FFT to 100 PCA components.\\
The best way to evaluate a compression procedure is to compare a compressed and afterwards reconstructed input to the original data. Figure \ref{fig:preprocessing_difference_waveform} compares exemplarily the recorded raw audio input of a music sample from the MusicNet dataset to a compressed and afterwards reconstructed version of the signal. The plot shows that the two signals are similar in their general behavior and only vary vary only in small details.\\
The same results can be observed by listening to the original and the compressed and reconstructed audio file: After preprocessing the audio contains some cracking noise, but nevertheless is the melody and the tone of the music well audible.\\
Summarizing, in neither by cutting after the first 150 FFT coefficients, nor by the PCA significant information of the original input is lost. Therefore, the described preprocessing leads to a decent compression of the audio signal.\\
\begin{figure}[htbp]
  \centering
  \includegraphics[width=\columnwidth]{figures/pca_explained_variance}
  \caption{Explained variance of the i'th eigenvector in the principal component analysis (with i being the horizontal axis)}
  \label{fig:pca_explained_variance}
\end{figure}

@@

Nevertheless, the experiments of this paper can be analyzed more detailed. Varying the stride and including normalization yielded conceptually different results.\\

@@

A second conceptual attempt during the experiments was to normalize the PCA results before the training. The aim of this additional preprocessing task was to balance the contribution of all 100 PCA dimensions to the loss, hoping that all frequencies are learned more equally. In a first attempt, data normalized to zero mean and unit variance turned out to produce too small loss for training. With vanishing gradients the network could not learn at all. Therefore, in a second attempt the input data was normalized to zero mean and an equal variance of all dimensions of 10. With this specifications, the network was able to learn as well as without normalization.\\

@@

\section{Discussion}
The approach of this paper could be promising due to the following reason: The original WaveNet paper's network has to learn the behaviour of sound consisting of repetitive oscillations. With transforming the signal to the Fourier space, we take this burden from the neural network. The network can focus on data which is more related to the final tone. For example while a tone is the same for a certain time, the output of the network is also the same.\\
But it might be exactly this phenomenon which makes good music generation impossible. During training, the network might do really well only staying on one and the same pitch, because changes happen in too rare occasions to influence the overall loss.\\
Future work based on that paper, one could try to include the rate of changing tones into the loss function and to punish a network when producing too many times in a row a similar output.\\
There are two reasons one could think of, why this approach's result does not match its expectations:
\begin{itemize}
  \item The original paper quantizes the amplitudes and can then use a softmax layer as a last layer of the network. In this new approach, we need continuous values (PCA coefficients). Since the results are much worse than with WaveNet, the missing quantization (and connected to it the missing softmax layer) might be the reason behind this phenomenon.
  \item The original WaveNet has a probability distribution as output. The exact amplitudes are then sampled according to that distribution. In the approach presented in this paper, one relies on continuous values, which are directly calculated (no probabilities included). In future work, one can change the model to produce a probability distribution of the output values.
\end{itemize}

It has been shown that the stride is of great importance for the strength of the frequencies. Strong frequencies are interpreted as the network being confident in its prediction while weak frequencies are a sign of uncertainty. When the training timesteps do not overlap (stride = window size), the network produces much weaker frequencies. But a big stride might also lead to what we experience as not-changing frequencies. With a small stride most of the data chunks are the same as the previous one (when the length of a tone is much longer than one time unit). With larger strides, the frequencies change abruptly from one window to another. When a change in frequencies occurs on an dataset with smaller strides, the absolute value of the difference of frequencies is much smaller.\\
Therefore one can assume that with changing the setting (stride, but also window size, dilations, normalization etc.), also this approach might produce good results.

\section{Summary}
With WaveNet a very promising music generation network is presented. It relies on predicting amplitudes when the preceding amplitudes are given. In this paper it is investigated whether transforming the data to the Fourier space improves the predictions. After transforming, the dimensionality of the audio file is vastly diminished. But at the same time, the ability to quantize the network's outputs is lost. Unfortunately the network could not return as good results as the original WaveNet paper. With the adopted network architecture, the generated audio files rapidly converge to one steady tone, while changes of pitches is not learned. However, by the approaches and discussions in the paper yields conceptual insights and motivate further investigations.
